---
title: "Exploring the Effects of Weather on Points Scored in NFL Games"
author: "Peter Christenson"
date: "12/27/2023"
output:
    pdf_document:
       latex_engine: xelatex
       toc: true
       number_sections: true
urlcolor: blue
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
---

```{r setup, include=FALSE}
################################################################################
### Title:  Analyze weather data
### Author: Peter Christenson 
################################################################################

rm(list = ls()); gc()
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  dplyr,
  lubridate,
  magrittr,
  readxl,
  writexl,
  survival,
  zoo,
  tidyr,
  stringr,
  janitor,
  readr,
  data.table,
  ggplot2,
  ggthemes,
  generics,
  knitr,
  kableExtra
)

##################################################
###  Globals
##################################################

# Directories
WORKING_DIRECTORY <- "/Users/peterchristenson/Desktop/Projects/NFL Weather Analysis/"
setwd(WORKING_DIRECTORY)
INPUT <- "./input/"
INTERMEDIATE <- "./intermediate/"
OUTPUT <- "./output/"

# Constants
DATE <- Sys.Date()
```

# Introduction

In this project, I explore the relationship between weather-related variables and scoring in NFL games played from the 2000-2022 seasons. I approach this in two ways:

* Conducting a hypothesis test to assess whether more points are scored in games played in indoor stadiums than in games played in non-indoor stadiums
* Constructing a regression to assess the relationship between weather-related variables and total points scored in games played outdoors

This analysis is largely exploratory and a more rigorous analysis would be necessary to fully assess the relationship between weather and scoring. However, the results of this analysis suggest that:

* The mean number of total points scored per game by both teams combined in games played indoors is greater than the mean number of total points scored per game by both teams combined in games played outdoors
* Weather by itself is a weak predictor for total scoring

```{r data, include=F}
##################################################
###  Data
##################################################

# Game data with weather
data_game_weather <- lapply(list.files(INTERMEDIATE, full.names = T), read.csv) %>% 
  bind_rows() %>% 
  clean_names() %>% 
  select(-x)

##################################################
###  Data cleaning
##################################################

#----- Create/clean relevant columns -----#
data_game_weather_cleaned <- data_game_weather %>% 
  # Create season variable (each season ends with Super Bowl)
  mutate(
    season = lag(cumsum(ifelse(grepl("Super Bowl", teams), 1, 0)), default = 0) + 2000 # First season
  ) %>% 
  select(season, everything()) %>% 
  # Clean other variables
  mutate(
    # Location
    location = gsub(".*Location: ", "", date),
    # Date
    date = gsub(" Location.*", "", date),
    date = gsub(".*Date: ", "", date),
    date = gsub(" ,", ", ", date),
    date = gsub("Febraury", "February", date),
    date = as.Date(date, format = "%B %d, %Y"),
    # Venue
    venue = gsub(" Attendance.*", "", venue),
    venue = gsub(".*Venue: ", "", venue),
    # Teams
    teams = gsub("<title>|</title>| Game Statistics| - Pro Football Archives", "", teams),
    teams = gsub(".* Championship - |.* Game - |.*Playoff - |.*Super Bowl [IVXLCDM]+ - |.*Super Bowl 50 - ", "", teams),
    teams = ifelse(
      season %in% c(2000:2007, 2013),
      gsub(" -.*", "", teams),
      gsub(".*- ", "", teams)
    ),
    teams  = gsub("Buccanneers", "Buccaneers", teams),
    team_1 = gsub(" at.*", "", teams),
    team_2 = gsub(".*at ", "", teams),
    # Weather
    weather_type = gsub(".*Weather: ", "", weather),
    weather_type = gsub(" Temp.*| [0-9].*", "", weather_type),
    weather_type = ifelse(startsWith(weather_type, "Temp"), NA, weather_type),
    temp_f       = gsub(".*Temp: ", "", weather),
    temp_f       = gsub(" F.*", "", temp_f),
    temp_f       = as.numeric(temp_f),
    wind_mph     = str_remove_all(tolower(gsub(".*Wind: ", "", weather)), " "),
    wind_mph     = as.numeric(gsub(".*?(\\d+)mph.*", "\\1", wind_mph)),
    humidity_pct = gsub(" wind.*", "", tolower(weather)),
    humidity_pct = as.numeric(gsub(".*: |%|,", "", humidity_pct))
  ) %>% 
  # Add relevant columns
  mutate(
    indoors   = ifelse(grepl("indoors", tolower(weather)), 1, 0),
    total_pts = team_1_pts + team_2_pts
  ) %>% 
  select(-c(teams, weather))

#----- Recode errors -----#
data_game_weather_cleaned %<>%
  mutate(
    temp_f = case_when(
      temp_f == 7777 ~ 77,
      T ~ temp_f
    )
  )
```

# Data

I web scraped boxscores of NFL games from the 2000 through 2022 seasons from [Pro Football Archives](https://www.profootballarchives.com), which provides scores for games as well as other characteristics such as location and relevant weather information.^[While I could not find any explicit statement that scraping from this website is a violation, I advise that any scraping done be done responsibly and with caution.] After some cleaning, I have a dataset with the scores of all games, the teams involved, the location, and weather information.

```{r view data, echo=FALSE, results = 'asis'}
# kable(head(data_game_weather_cleaned), format = "latex") %>% 
#    kable_styling(latex_options = "scale_down")
kable(head(data_game_weather_cleaned), format = "latex", caption = "Cleaned Web Scraped Data") %>% 
  kable_styling(latex_options = c("HOLD_position","scale_down"))
```
Note that the above table is just 5 rows of the cleaned data.

# Analysis
## Hypothesis Testing
I was first interested in exploring the difference in scoring between games played indoors versus games played outdoors. Specifically, I was interested in conducting a two-sample t-test comparing the mean number of points scored in both circumstances.
```{r plot data,echo=F, out.width="75%", fig.align = 'center', fig.cap="Kernel Density Plot of Total Points Scored by Stadium Type"}
#----- Create dataframe for test -----#
indoor_outdoor <- data_game_weather_cleaned %>% 
  select(indoors, total_pts) %>% 
  mutate(
    indoors = as.character(indoors),
    indoors = recode(
      indoors,
      '1' = 'Yes',
      '0' = 'No'
    )
  )

#----- Plot -----#
plt_indoors_outdoors <- ggplot(indoor_outdoor) + 
  geom_density(aes(total_pts, fill = indoors), alpha = 0.3) +
  theme_few() +
  guides(fill=guide_legend(title="Indoor Stadium")) +
  theme(plot.caption = element_text(hjust = 0)) +
  labs(
    # title   = "Total Points Scored by Stadium Type",
    x       = "Total Points in Game"
    # caption = "Notes:\n[1] Games played from the 2000-2022 seasons." 
  )
plot(plt_indoors_outdoors)
```
Before I could perform the t-test, I checked:

1. Whether both groups followed a normal distribution
2. Whether there is homogeneity of variance

As both groups have a large amount of observations, I assume the Central Limit Theory holds and that both groups are normally distributed.
``` {r group size, echo=F, results = 'asis'}
num_obs <- indoor_outdoor %>% 
  count(indoors) %>% 
  rename(num_games = n)
kable(num_obs, format = "latex", caption = "Number of Games by Stadium Type") %>% 
  kable_styling(latex_options = c("HOLD_position"))
```
To compare the variances of both groups, I conduct an F test.
```{r variance test, echo=F}
points_indoor <- indoor_outdoor$total_pts[indoor_outdoor$indoors == "Yes"]
points_non_indoor <- indoor_outdoor$total_pts[indoor_outdoor$indoors == "No"]

var.test(points_indoor, points_non_indoor)
```
Per the results of this test, I assume the variances are equal due to the large p-value.

Since both conditions are satisfied, I can move forward with the t-test. Since indoor games provide a more controlled football-playing atmosphere (temperature, wind, precipitation, etc.), my hypothesis was that games played indoors have more points scored. Let group 1 represent games played indoors and group 2 represent games played outdoors, and $\mu$ represent the mean total number of points scored per game.

&emsp;&emsp; \(H₀\): $\mu_{1}$ $\le$ $\mu_{2}$

&emsp;&emsp; \(Hₐ\): $\mu_{1}$ > $\mu_{2}$

``` {r t-test, echo=F}
t.test(points_indoor, points_non_indoor, alternative = "greater", conf.level = 0.95)
```
As the p-value is near zero, I reject the null hypothesis and conclude that the mean total number of points scored per game is greater in games played indoors than in games played outdoors.

## Regression

I also performed regression analysis to assess the effects of weather on scoring. The first model I constructed considered only variables that pertain to weather as independent variables, specifically rain, snow, temperature, wind, and percent humidity. My dependent variable is total points scored in the game. The model is then as follows:

\[\text{Total Points Scored} = \beta_0 + \beta_1D_{rain} + \beta_2D_{snow} + \beta_3Temperature + \beta_4Wind + \beta_5Humidity + \varepsilon \]

Where \(D_{rain}\) and \(D_{snow}\) refer to binary 'dummy' variables that indicate the presence of rain or snow, and \(Temperature\), \(Wind\), and \(Humidity\) referring to the temperature (in degrees Fahrenheit), the wind speed (in mph), and percent humidity, respectively.
```{r reg1, echo=F}
#----- Regression prep -----#
regression_df <- data_game_weather_cleaned %>% 
  mutate(rain   = ifelse(grepl("rain|showers|drizzle", tolower(weather_type)), '1', '0'),
         snow   = ifelse(grepl("snow", tolower(weather_type)), '1', '0'),
         season = as.character(season)) %>% 
  # Only consider games with weather data
  filter(!(is.na(temp_f) | is.na(wind_mph) | is.na(humidity_pct)),
         indoors == 0)

reg_variables <- c("season", "location", "temp_f", "team_1", "team_2", 
                   "wind_mph", "humidity_pct", "rain", "snow", "total_pts")

regression_df %<>%
  select(all_of(reg_variables))

#----- Regression -----#
reg1 <- lm(total_pts ~ rain + snow + temp_f + wind_mph + humidity_pct,
          data = regression_df)

summary(reg1)
```
These results suggest weather by itself is a fairly weak predictor for the total amounts of points scored in a game. 

To improve the regression, I introduced new variables: the location of the game, and the two teams involved.^[I consider the interaction between the season of the game and two teams. For example, if a game took place in the 2017 season between the Packers and the Vikings, I would consider the 2017 Packers and 2017 Vikings as independent variables (as opposed to the teams by themselves without any season).] The new model is then constructed as:

\[
\begin{aligned}
\text{Total Points Scored} &= \beta_0 + \beta_1D_{rain} + \beta_2D_{snow} + \beta_3\text{Temperature} + \beta_4\text{Wind} \\
&\quad + \beta_5\text{Humidity} + \beta_6\text{Location} + \beta_7\text{Team}_1 + \beta_8\text{Team}_2 + \varepsilon 
\end{aligned}
\]

The results of which are below.
```{r reg2, echo=F}
reg <- lm(total_pts ~ (season*team_1) + (season*team_2) + location + rain + snow + temp_f + wind_mph + humidity_pct,
          data = regression_df)

reg_summ <- summary(reg)
cat(paste0("Multiple R-squared: ", round(reg_summ$r.squared, 4), "\nAdjusted R-squared: ", round(reg_summ$adj.r.squared, 4)))

```
Comparing the \(R^2\) values across both models, the inclusion of location and the teams playing improves the fit of the model considerably.
